---
title: "Building intuition around probability distributions"
author: "BRG"
date: "2023-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r, message=F}
set.seed(1999)
library(tidyverse)
```


# Discrete distributions

### Bernoulli distribution

Outcome is 1 with probability $p$ and 0 with probability $1-p$.

In R, to simulate Bernoulli variables, we use the binomial distribution functions
with parameter size=1.

```{r}
# Simulate one random Bernoulli variable with probability 0.6
rbinom(n = 1, size = 1, prob = 0.6)

# Simulate 1000 random Bernoulli variables with probability 0.6
bern_vars <- rbinom(n = 1000, size = 1, prob = 0.6)

# How many of those were 1?
table(bern_vars)

# What was the mean response? The variance?
# We expect: mean = 0.6, var = 0.6(1-0.6) = 0.24
mean(bern_vars); var(bern_vars)

```

### Binomial distribution

How many of $N$ trials are successful, if each trial is successful with probability $p$?


```{r}
# Simulate one random binomial variable with probability 0.6 and size 10
rbinom(n = 1, size = 10, prob = 0.6)

# Simulate 1000 random Bernoulli variables with probability 0.6 and size 10
binom_vars <- rbinom(n = 1000, size = 10, prob = 0.6)

# Look at the distribution, compared to the expected distribution.
x <- 0:10
expectation <- dbinom(x, size = 10, prob = 0.6)

ggplot() +
  geom_histogram(data=data.frame(var = binom_vars), aes(var)) +
  geom_line(data = data.frame(x = x, expectation = expectation*1000),
           aes(x, expectation), inherit.aes = FALSE) +
  ggtitle("Histogram of binomial trials compared to expectation (line)")


# What was the mean response? The variance?
# We expect: mean = 0.6, var = 0.6(1-0.6) = 0.24
mean(bern_vars); var(bern_vars)

```

### Poisson distribution

The Poisson arises from a spatial point process.

Consider a case where we have 100 individuals who are at any given time distributed randomly over a 10x10 area. If we put down a 1x1 quadrat and counted the number of individuals in one cell, the count that we get follows a Poisson distribution.


To demonstrate this let's start by simulating the positions of 100 individuals 
randomly in 2D space.

```{r}
# 100 individuals

gen_rand_points <- function(n, spatial_dim) {
  # runif is a function to generate a random number from a uniform distribution.
  # In this case, our number can be anything between 0-10.
  dat <- data.frame(x = runif(n, 0, spatial_dim), y = runif(n, 0, spatial_dim))
  
  return(dat)
}

plot_rand_points <- function(dat, spatial_dim, resoln) {
  grid <- data.frame(x = seq(0, spatial_dim, by = resoln))
  
  ggplot(dat, aes(x, y)) + 
    geom_vline(data = grid, aes(xintercept = x), col = "darkgray") +
    geom_hline(data = grid, aes(yintercept = x), col = "darkgray") +
    geom_vline(xintercept = c(0, spatial_dim), col = "black", linewidth = 2) +
    geom_hline(yintercept = c(0, spatial_dim), col = "black", linewidth = 2) +
    geom_point(col = "darkred", size = 2) + theme_minimal() +
    theme(panel.grid = element_blank()) +
    ylim(0, spatial_dim) + xlim(0, spatial_dim)
    
}

count_rand_points <- function(dat, spatial_dim, resoln = 1) {
  dat$xgrid <- floor(dat$x / resoln)
  dat$ygrid <- floor(dat$y / resoln)
  dat$cell <- dat$xgrid + (spatial_dim/resoln)*(dat$ygrid)
  
  counts <- data.frame(cell = 1:(spatial_dim/resoln)^2) %>% 
    left_join(count(dat, cell), by = "cell") %>% 
    mutate(n = ifelse(is.na(n), 0, n))
  
  return(counts)
}



# Simulate 200 individuals on a 10x10 grid
n <- 200
spatial_dim <- 10
grid_resoln <- 2

sim_locs <- gen_rand_points(n, spatial_dim)
plot_rand_points(sim_locs, spatial_dim, grid_resoln)

cell_counts <- count_rand_points(sim_locs, spatial_dim, grid_resoln)

# Counts follow a Poisson distribution where the expected value is
# (num. individuals) * (cell area) / (total area)
x <- 0:(max(cell_counts$n) + 3)
expectation <- dpois(x, n * (grid_resoln^2) / (spatial_dim^2))

ggplot() +
  geom_col(data=count(cell_counts, n), aes(n, nn)) +
  geom_point(data = data.frame(x = x, expectation = expectation*nrow(cell_counts)),
           aes(x, expectation), inherit.aes = FALSE, col = "red", size = 3) +
  ggtitle("Histogram of cell counts trials compared to Poisson expectation (points)")
```
This relationship is approximate. Why? Where does this break down? (Think about the domain of the Poisson distribution.)

### Takeaways

- For a given distribution family `dist`, you can use `rdist()` to simulate random
draws given parameters, and you can use `ddist()` to get the probability mass or
probability density at a certain value or values given parameters.

- The Poisson distribution describes a process corresponding to the random arrangement of individuals in 2D space. 







